### 2017-07-07 isolation.tex
Edits.

A distinctive characteristic of our development is that it is
parameterized on the isolation specification $\I$. $\I$ can be
instantiated with the declarative characterization of an isolation
guarantee or a concurrency control mechanism, regardless of its actual
implementation. This allows us to model a range of isolation
properties that are relevant to the theory and practice of realistic
transaction processing systems, some of the most well-known of which are
given below:

@@

\textbf{Unique Ids}. As the \C{new\_order} example
(Sec.~\ref{sec:motivation}) demonstrates, enforcing global uniqueness
of ordered identifiers requires stronger isolation than possible using
a weak read committed isolation level. Alternatively, globally unique
sequence numbers, regardless of the isolation level, can be requested
from a relational databases via SQL's \C{UNIQUE} and
\C{AUTO\_INCREMENT} keywords. Our development crucially relies on the
uniqueness of record identifiers\footnote{This has parallels in
  real-world. For e.g., MySQL's InnoDB engine automatically adds a
  6-byte unique identifier if none exists for a record.}, which are
generated locally by the \rulelabel{E-Insert} rule.  The global
uniqueness of locally generated identifiers can be axiomatized as an
isolation property as following:
\begin{smathpar}
\begin{array}{lcl}
  \I_{id}(\stl,\stg,\stg') & = & \forall(r\in\stl).~
      r.\idf\notin \dom(\stg) \Rightarrow r.\idf\notin \dom(\stg').
\end{array}
\end{smathpar}
$\I_{id}$ ensures that a globally unique id generated during a
transaction's execution remains globally unique until it commits,
prohibiting interference from a concurrent transaction that adds the
same id. The axiom thus simulates a global counter protected by an
exclusive lock without explicitly appealing to specific
implementation artifacts.

@@

\textbf{Snapshots} Almost all major relational databases implement
isolation levels that execute transactions against a static snapshot
of the database that can be axiomatized thus:
\begin{smathpar}
\begin{array}{lcl}
  \I_{ss}(\stl,\stg,\stg') & = & \stg' = \stg.
\end{array}
\end{smathpar}
% The snapshot semantics however may not be applicable throughout the
% lifetime of the transaction, like in the case of Repeatable Read (RR).

\textbf{Read-Only Transactions}. Certain databases implement special
privileges for read-only transactions. Read-only behavior can be
enforced on a transaction by including the following proposition as
part of its isolation invariant:
\begin{smathpar}
\begin{array}{lcl}
  \I_{ro}(\stl,\stg,\stg') & = & \stl = \emptyset\\
\end{array}
\end{smathpar}
%% If a transaction declared as read-only performs a write, then its
%% $\stl\neq \emptyset$, and the transaction never commits.

In addition to these properties, there are various isolation
guarantees proposed in the database or distributed systems literature,
or implemented by commercial vendors that can also be specified within
this framework:

@@

\textbf{Repeatable Read (RR)} By definition, a read to a transactional
variable in a Repeatable Read transaction is required to return the
same value whenever accessed.  RR is often implemented (for e.g., in
~\cite{mysqliso,bailishat}) by executing the transaction against a
(conceptual) snapshot of the database, but committing its writes to
the actual database. This implementation of RR can be axiomatized as
$\I_e = \I_{ss}$ and $\I_{c}=true$. However, this specification of RR
is stronger than the ANSI SQL specification, which requires no more
than the invariance of already read records. In particular, ANSI SQL
RR allows \emph{phantom reads}, a phenonmenon in which a repeated
\C{SELECT} query might return newly inserted records that were not
previously returned. This specification is implemented, for e.g., in
Microsoft's SQL server, using record-level exclusive read locks, that
prevent a record from being modified while it is read by an
uncommitted transaction.  The ANSI SQL RR specification can be
axiomatized in our framework, but it requires a minor extension to our
operational semantics to track a transaction's reads. In particular,
the records returned by \C{SELECT} should be added to the local
database $\stl$, but without changing their transaction identifiers
($\txnf$ fields), and flush ($\gg$) should only flush the records that
bear the current transaction's identifier. With this extension, ANSI
SQL RR can be axiomatized thus:
\begin{smathpar}
\begin{array}{lcl}
  \I_e(\stl,\stg,\stg') & \Leftrightarrow & \forall(r\in\stl).
      r \in \Delta \Rightarrow r \in \Delta'\\
  \I_c(\stl,\stg,\stg') & \Leftrightarrow & true\\
\end{array}
\end{smathpar}
If a record $r$ belongs to both $\stl$ and $\stg$, then it must be a
record written by a different transaction and read by the current
transaction. By requiring $r\in\stg'$, $\I_e$ guarantees the
invariance of $r$, thus the repeatability of the read.

\textbf{Snapshot Isolation (SI)} The concept of executing a
transaction against a consistent snapshot of the database was first
proposed as Snapshot Isolation in~\cite{berenson}. SI doesn't admit
write-write conflicts, and the original proposal, which is implemented
in Microsoft SQL Server, required the database to roll-back an SI
transaction if conflicts are detected during the commit. This behavior
can be axiomatized thus:
\begin{smathpar}
\begin{array}{lcl}
\I_e\,\,(\stl,\stg,\stg') & = & \stg' = \stg\\
\I_c\,\,(\stl,\stg,\stg') & = & \forall(r\in\stl)(r'\in\stg).~ r'.\idf = r.\idf \Rightarrow r'\in\stg'.
\end{array}
\end{smathpar}
Observe that the same axiomatization applies to Postgres's RR,
although its implementation (described in Sec.~\ref{sec:motivation})
differs considerably from the original proposal. Thus, reasoning done
for an SI transaction on MS SQL server carries over to Postgres's RR
and vice-versa, demonstrating the benefits of reasoning axiomatically
about isolation properties.

>>>>

A distinctive characteristic of our development is that it is
parameterized on the isolation specification $\I$. $\I$ can be
instantiated with the declarative characterization of an isolation
guarantee or a concurrency control mechanism, regardless of its actual
implementation. This allows us to model a range of isolation
properties that are relevant to the theory and practice of transaction
processing systems. Few such well-known properties are discussed
below:

@@

\textbf{Unique Ids}. As the \C{new\_order} example
(Sec.~\ref{sec:motivation}) demonstrates, enforcing global uniqueness
of ordered identifiers requires stronger isolation levels than ones
that are default on most databases. Alternatively, globally unique
sequence numbers, regardless of the isolation level, can be requested
from a relational databases via SQL's \C{UNIQUE} and
\C{AUTO\_INCREMENT} keywords. Our development crucially relies on the
uniqueness of record identifiers\footnote{This has parallels in
  real-world. For e.g., MySQL's InnoDB engine automatically adds a
  6-byte unique identifier if none exists for a record.}, which are
checked locally for uniqueness by the \rulelabel{E-Insert} rule. The
global uniqueness of locally unique identifiers can be axiomatized as
an isolation property as following:
\begin{smathpar}
\begin{array}{lcl}
  \I_{id}(\stl,\stg,\stg') & = & \forall(r\in\stl).~
      r.\idf\notin \dom(\stg) \Rightarrow r.\idf\notin \dom(\stg').
\end{array}
\end{smathpar}
$\I_{id}$ ensures that if the id of a record is globally unique when
it is added to a transaction's $\stl$, it remains globally unique
until the transaction commits. This is achieved by prohibiting the
interference from a concurrent transaction that adds the same id. The
axiom thus simulates a global counter protected by an exclusive lock
without explicitly appealing to an implementation artifact.

@@

\textbf{Snapshots} Almost all major relational databases implement
isolation levels that execute transactions against a static snapshot
of the database that can be axiomatized thus:
\begin{smathpar}
\begin{array}{lcl}
  \I_{ss}(\stl,\stg,\stg') & = & \stg' = \stg.
\end{array}
\end{smathpar}

\textbf{Read-Only Transactions}. Certain databases implement special
privileges for read-only transactions. Read-only behavior can be
enforced on a transaction by including the following proposition as
part of its isolation invariant:
\begin{smathpar}
\begin{array}{lcl}
  \I_{ro}(\stl,\stg,\stg') & = & \stl = \emptyset\\
\end{array}
\end{smathpar}

In addition to these properties, various specific isolation levels
proposed in the database or distributed systems literature, or
implemented by commercial vendors can also be specified within this
framework:

@@

\textbf{Repeatable Read (RR)} By definition, multiple reads to a
transactional variable in a Repeatable Read transaction are required
to return the same value.  RR is often implemented (for e.g., in
~\cite{mysqliso,bailishat}) by executing the transaction against a
(conceptual) snapshot of the database, but committing its writes to
the actual database. This implementation of RR can be axiomatized as
$\I_e = \I_{ss}$ and $\I_{c}=true$. However, this specification of RR
is stronger than the ANSI SQL specification, which requires no more
than the invariance of already read records. In particular, ANSI SQL
RR allows \emph{phantom reads}, a phenonmenon in which a repeated
\C{SELECT} query might return newly inserted records that were not
previously returned. This specification is implemented, for e.g., in
Microsoft's SQL server, using record-level exclusive read locks, that
prevent a record from being modified while it is read by an
uncommitted transaction. The ANSI SQL RR specification can be
axiomatized in our framework, but it requires a minor extension to our
operational semantics to track a transaction's reads. In particular,
the records returned by \C{SELECT} should be added to the local
database $\stl$, but without changing their transaction identifiers
($\txnf$ fields), and flush ($\gg$) should only flush the records that
bear the current transaction's identifier. With this extension, ANSI
SQL RR can be axiomatized thus:
\begin{smathpar}
\begin{array}{lcl}
  \I_e(\stl,\stg,\stg') & \Leftrightarrow & \forall(r\in\stl).
      r \in \Delta \Rightarrow r \in \Delta'\\
  \I_c(\stl,\stg,\stg') & \Leftrightarrow & true\\
\end{array}
\end{smathpar}
If a record $r$ belongs to both $\stl$ and $\stg$, then it must be a
record written by a different transaction and read by the current
transaction (since the current transaction's records are not yet
present in $\stg$). By requiring $r\in\stg'$, $\I_e$ guarantees the
invariance of $r$, thus the repeatability of the read.

\textbf{Snapshot Isolation (SI)} The concept of executing a
transaction against a consistent snapshot of the database was first
proposed as Snapshot Isolation in~\cite{berenson}. SI doesn't admit
write-write conflicts, and the original proposal, which is implemented
in Microsoft SQL Server, required the database to roll-back an SI
transaction if conflicts are detected during the commit. This behavior
can be axiomatized as $\I_e = \I_{ss}$ (execution against a snapshot),
and $\I_c = \I_{ww}$ (avoiding write-write conflicts during the
commit).
% \begin{smathpar}
% \begin{array}{lcl}
% \I_e\,\,(\stl,\stg,\stg') & = & \stg' = \stg\\
% \I_c\,\,(\stl,\stg,\stg') & = & \forall(r\in\stl)(r'\in\stg).~ r'.\idf = r.\idf \Rightarrow r'\in\stg'.
% \end{array}
% \end{smathpar}
Note that the same axiomatization applies to Postgres's RR,
although its implementation (described in Sec.~\ref{sec:motivation})
differs considerably from the original proposal. Thus, reasoning done
for an SI transaction on MS SQL server carries over to Postgres's RR
and vice-versa, demonstrating the benefits of reasoning axiomatically
about isolation properties.
